---
title: "Introductory exercises with bioRad"
author: Adriaan M. Dokter
output: bookdown::gitbook
---

```{r setup, echo=FALSE, message=FALSE}
SHOW_ANSWERS <- FALSE
if (Sys.info()["sysname"] == "Linux") prefix <- "/home/adriaan/" else prefix <- "/Users/amd427/"
if (SHOW_ANSWERS) knitr::opts_knit$set(root.dir = normalizePath(paste(prefix, "git/training/amsterdam2026/", sep = "")))
# knitr::opts_chunk$set(eval=FALSE)
Sys.setenv(TZ = "UTC")
Sys.setenv(PROJ_LIB = "/opt/homebrew/Cellar/proj/9.7.1/share/proj/")
library(bioRad)
library(terra)
library(sf)
library(dplyr)
```

# Getting started

To install the latest bioRad version follow these [install instructions](https://adriaandokter.com/delaware2024/bioRad_installation.html)

Execute each of the code examples provided below in [RStudio](https://www.rstudio.com/), and try to complete the exercises.

```{r, eval=SHOW_ANSWERS}
# make sure you start with a fresh R session
# load the bioRad package
library(bioRad)
# load the getRad package, which we will use to access radar data:
library(getRad)
# some other useful (spatial) packages
library(terra)
library(sf)
library(dplyr)
# check the package versions
packageVersion("bioRad") # should be 0.12.0.9000
packageVersion("vol2birdR") # should be 1.3.0
packageVersion("getRad") # should be 0.3.0
```

All bioRad's functions are documented in an extensive [function reference](https://adriaandokter.com/bioRad/reference/index.html) online, as well as in manual pages within R:

```{r, eval=FALSE}
# bring up the package general help page:
?bioRad
```

Start by making a new directory on your local machine that you will use for this practical:

```{r, eval=FALSE}
# make a new local directory on your machine for this practical
# replace the string below with the path of that directory:
HOME <- "your/personal/working/directory/"
# check that the directory exists. If the next statement evaluates to FALSE, something went wrong: the directory does not exist or you didn't specify its path correctly
file.exists(HOME)
# we will make HOME our work directory, the default folder in which R will look
# for files, and where it will output newly generated files.
setwd(HOME)
# Finally, we set the local time zone to UTC, so all plotted time axes will be in UTC
Sys.setenv(TZ = "UTC")
```

Your R session is now properly set up.

# Basic visualization of radar scans

## The structure of polar volumes

```{r, eval=SHOW_ANSWERS, results='hide'}
# Let's load a NEXRAD polar volume file for the KHGX radar (Houston):
my_pvol <- get_pvol("KHGX", as.POSIXct("2017-05-04 01:25:00"))
```

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS}
# The default summary information of a `pvol` object contains information
# on the scans (sweeps) and their moments:
my_pvol
```

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS}
# We can also extract the elevation angles from the polar volume as follows:
get_elevation_angles(my_pvol)
```

## Plotting radar scans

```{r, eval=SHOW_ANSWERS, warning=FALSE}
# let's extract the scan collected at 0.5 degree elevation from our polar volume:
my_scan <- get_scan(my_pvol, 0.5)
# print some information about this scan:
my_scan
# let's plot the reflectivity factor parameter of the scan in a range - azimuth coordinate system:
plot(my_scan, param = "DBZH")
```

**Exercise 1:** Plot the radial velocity parameter (VRADH) of the 1.5 degree elevation scan in a range - azimuth coordinate system. (See manual page of the `read_pvolfile()` function for the nomenclature of various available quantities).

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# Plot the radial velocity (VRADH) quantity of the 1.5 degree elevation scan:
plot(get_scan(my_pvol, 1.5), param = "VRADH")
```

Often it is easier to visually explore radar scans as a PPI (plan position indicator), which is a projection of the scan on a Cartesian (X,Y) or (lat,lon) grid:
```{r, eval=SHOW_ANSWERS, warning=FALSE}
# before we can plot the scan, we need to project it on a Cartesian grid,
# i.e. we need to make a Plan Position Indicator (PPI)
my_ppi <- project_as_ppi(my_scan)
# print some information about this ppi:
my_ppi
# you can see we projected it on a 500 meter grid
# (check the manual of the project_as_ppi function to see how you can
# change the grid size (argument grid_size) and the maximum distance
# from the radar up to where to plot data (argument range_max))
#
# Now we are ready to plot the ppi, for example let's plot reflectivity factor DBZH:
plot(my_ppi, param = "DBZH")
```

**Exercise 2:** This case shows an incoming precipitation front, characterized by localized but intense thunderstorms, as well as biological scattering. Make also a ppi plot of the correlation coefficient (RHOHV) and radial velocity (VRADH). Verify which regions are precipitation, and infer from those patterns the (approximate) direction of movement of biology and precipitation.

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# Plot the correlation coefficient (RHOHV):
plot(my_ppi, param = "RHOHV")
```

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# Plot the radial velocity (VRADH):
plot(my_ppi, param = "VRADH")
```

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# Answer:
# * Precipitation areas are characterized by high correlation coefficients (typically > 0.95),
#   i.e. the top left corner of the image is precipitation
# * The blue radial velocity of the precipitation indicates it is moving towards the radar.
# * The radial velocity field of the biology shows areas south-west of the radar moving
#   towards the radar (blue), and areas north-east of the radar moving away from it (red).
#   The biology is therefore moving towards the north-east
```

**Exercise 3:** Based on the radial velocity image, are the biological scatterers birds or insects? Why?

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# Answer:
# * Precipitation moves with the wind field. Since the biology moves into a different direction
#   than the wind and a very different speed, we can be sure these are birds.
#   The biological scatterers have a high self-propelled speed, which is
#   typical for birds, not for insects.
```

## Overlaying radar scans on maps

```{r, eval=SHOW_ANSWERS, warning=FALSE}
# It is often informative to plot radar data on a base layer map.
# Function `bioRad::map`  overlays a PPI on a basemap (we restrict the color scale from -20 to 40 dBZ):
map(my_ppi, param = "DBZH", zlim = c(-20, 40), alpha=0.4)
```

# Screening out weather

## using correlation coefficient

Screening precipitation based on correlation coefficient is arguably the most simple and established approach for screening out weather in cases where dual-polarization data is available. It is based on removing data above a certain $\rho_{/HV}$ tresholds, most commonly 0.95.
```{r, eval=SHOW_ANSWERS}
# We will screen out the reflectivity areas with for which the
# correlation coefficient is > 0.95 (indicating precipitation)
# We do so by redefining parameter DBZH, replacing pixels with NA values
# whenever RHOHV > 0.95
my_ppi_clean <- calculate_param(my_ppi, DBZH = ifelse(RHOHV > 0.95, NA, DBZH))
# plot the original and cleaned up reflectivity:
map(my_ppi_clean, param = "DBZH", zlim = c(-20, 40), map="osm")
```

## using MistNet
You can also use MistNet for screening out weather when
* the radar operates at S-band wavelengths
* you only have single polarization data available (specifically, the three basic quantities radial velocity VRADH, reflectivity DBZH, and spectrum width WRADH).

```{r, eval=SHOW_ANSWERS}
# apply the MistNet model to the polar (pvol):
my_pvol <- apply_mistnet(my_pvol)
# mistnet will add additional parameters to the
# elevation scans at 0.5, 1.5, 2.5, 3.5 and 4.5 degrees
# let's extract the scan closest to 0.5 degrees:
my_scan <- get_scan(my_pvol, 0.5)
# plot some summary info about the scan to the console:
my_scan
```

MistNet adds several new parameters:

* `WEATHER`: a probability (0-1) for the weather class
* `BIOLOGY`: a probability (0-1) for the biology class
* `BACKGROUND`: a probability (0-1) for the background (empty space) class
* `CELL`: the final segmentation, which is based on the WEATHER parameters of all five MistNet elevation scans. `CELL` also includes a border calculated by a region growing approach. `CELL` values equal 1 for the additional border, and 2 for the originally segmented precipitation area by MistNet.

```{r, eval=SHOW_ANSWERS}
# as before, project the scan as a ppi:
my_ppi <- project_as_ppi(my_scan)
# plot the probability for the WEATHER class
plot(my_ppi, param = 'WEATHER')
# plot the final segmentation result:
# plot the probability for the WEATHER class
plot(my_ppi, param = 'CELL')
# let's remove the identified precipitation area (with additional 5 km border) from the ppi, and plot it:
my_ppi_clean_mistnet <- calculate_param(my_ppi, DBZH = ifelse(CELL >= 1, NA, DBZH))
map(my_ppi_clean_mistnet, param = 'DBZH')
# let's remove the identified precipitation area (without additional border) from the ppi:
my_ppi_clean_mistnet <- calculate_param(my_ppi, DBZH = ifelse(CELL == 2, NA, DBZH))
map(my_ppi_clean_mistnet, param = 'DBZH')
```

**Exercise 4:** Examine the differences in weather screening by MistNet and by correlation coefficient thresholding by looking at the radial velocity field. Can you identify a shortcoming of each method for this particular case?

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS}
# Set VRADH to NA when correlation coefficient is > 0.95
my_ppi_clean <- calculate_param(my_ppi, VRADH = ifelse(RHOHV > 0.95, NA, VRADH))
# The VRADH plot shows that birds and weather have different directions
# Weather is southbound coming in at the northern end of the image
# Birds are moving towards the north east.
# RHOHV thresholding doesn't fully remove the incoming precipitation
# at the southern edge of the incoming weather
plot(my_ppi_clean, param="VRADH")
# Set VRADH to NA when MistNet identifies precipitation (excluding 5 km border)
my_ppi_clean_mistnet <- calculate_param(my_ppi, VRADH = ifelse(CELL == 2, NA, VRADH))
# The VRADH plot shows that MistNet precipitation filtering does a better job in this
# case in removing the southbound weather.
# However it removes some areas that are likely birds in the center of the image.
plot(my_ppi_clean_mistnet, param="VRADH")
```

# Vertical profiles

In this section you will learn to compute, interpret and analyze vertical profiles (vp's). A vp consists of the (bird) density, speed and directions at different altitudes at the location of a single radar. It is typically calculated for all data within a cylinder of 35 km radius around the radar, i.e. only containing data at relatively short distances, where the radar beam is still sufficiently narrow to resolve altitude information.

Section 6 has examples that show how to process polar volume data into vertical profiles. To save time, we will start below with a list of pre-processed vertical profiles for the Brownsville radar in Texas (KBRO).

## Loading processed vertical profiles

```{r, eval=SHOW_ANSWERS}
# Pre-processed profiles are available for many radars, see getRad documentation for details
# We will download a day of pre-processed profiles from BirdCast:
# A time series of profiles we will refer to as `VPTS` data (Vertical Profile Time Series).
my_vpts <- get_vpts("KBRO", as.Date("2017-05-14"), source="birdcast")
# print some info for the vpts object. It should contain 257 profiles
my_vpts
```

## Inspecting single vertical profiles

Now that you have loaded a VPTS, we can start exploring it. We will start with plotting and inspecting single vertical profiles, i.e. a single profile from the time series.

```{r, eval=SHOW_ANSWERS}
# let's extract a profile from the list, in this example the 41st profile:
my_vp <- my_vpts[41]
# print some info for this profile to the console
my_vp
# test whether this profile was collected at day time:
check_night(my_vp)
# plot the vertical profile, in terms of reflectivity factor
plot(my_vp, quantity = "dbz")
# plot the vertical profile, in terms of (linear) reflectivity
plot(my_vp, quantity = "eta")
```

`eta` and `dbz` are closely related, the main difference is that reflectivity factors are logarithmic, and reflectivities linear. You can convert one into the other using `eta_to_dbz()` and `dbz_to_eta()` functions, which follows this simple formula:

```
eta = (radar-wavelength dependent constant) * 10^(dbz/10)
```

The reflectivity factor `dBZ` is the quantity used by most meteorologist. It has the useful property that at different radar wavelengths (e.g. S-band versus C-band) the same amount of precipitation shows up at similar reflectivity factors. The same holds for insects, as well as any other target that is much smaller than the radar wavelength (S-band = 10 cm, C-band = 5 cm), the so-called Rayleigh-scattering limit.

In the case of birds we are outside the Rayleigh limit, because birds are of similar size as the radar wavelength. In this limit reflectivity `eta` is more similar between S-band and C-band. `eta` is also more directly related to the density of birds, since `eta` can be expressed as (bird density) x (radar cross section per bird). For these two reasons, for weather radar ornithologists reflectivity `eta` is the more conventional unit.

```{r, eval=SHOW_ANSWERS}
# let's plot the vertical profile, in terms of bird density
plot(my_vp, quantity = "dens")
# print the currently assumed radar cross section (RCS) per bird:
rcs(my_vp)
```

**Exercise 5:** If you change your assumption on the bird's radar cross section in the previous example, and assume the RCS is 10 times as large, what will be the effect on the bird density profile?

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# Answer:
# All bird densities will be a factor 10 times lower.
```

The assumed radar cross section can be changed as follows:

```{r, eval=SHOW_ANSWERS}
# let's change the RCS to 110 cm^2
rcs(my_vp) <- 110
```

**Exercise 6:** Verify your answers on the previous two questions, by re-plotting the vertical profiles for the bird density quantity.

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# After changing the RCS above, we simply plot the vertical profile again
plot(my_vp)
# Verify the densities are scaled down by a factor 10
```

## Plotting vertical profile time series

We will now examine the full Vertical Profile Time series we downloaded earlier, e.g. the vertical profiles obtained from a single radar over a full day.
```{r, eval=SHOW_ANSWERS}
# time series objects can be subsetted, just as you may be used to with vectors
# here we subset the first 50 timesteps:
my_vpts[1:50]
# here we extract a single timestep, which gives you back a vertical profile class object:
my_vpts[100]
# to plot the full time series from 0-3000 m above ground level:
plot(my_vpts[-c(40,80)], ylim=c(0,3000))
# remove one corrupted profile (timestep 52):
my_vpts <- my_vpts[-52]
# check the help file for the plotting function of profile time series
# Because profile timeseries are of class 'vpts', it's associated plotting function
# is called plot.vpts:
?plot.vpts
```

Let's make a plot for a subselection (night only) of the time series:

```{r, eval=SHOW_ANSWERS}
# filter our vpts for night time
my_vpts_night <- filter_vpts(my_vpts, night=TRUE)
# plot this smaller time series:
plot(my_vpts_night, ylim=c(0,3000))
```

**Exercise 7:** Interpret the wind barbs in the profile time series figure: what is the approximate speed and direction at 1500 meter at 6 UTC? In the speed barbs, each half flag represents 2.5 m/s, each full flag 5 m/s, [each pennant (triangle) 25 m/s, not occurring in this case].

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# Answer:
#
# At 1500 meter 6 UTC the wind barbs have 2 full flags and one half flag.
# Therefore the ground speed is approximately 2x5 + 2.5 = 12.5 m/s.
```

**Exercise 8:** Extract the vertical profile at 6 UTC from the time series and plot the vertical profile of ground speed (quantity `ff`). Hint: use the `nearest` argument of function `filter_vpts()` to extract the 6 UTC profile. Check whether your answer to the previous question was approximately correct.

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# First extract the profile at 6 UTC:
vp_6UTC <- filter_vpts(my_vpts_night, nearest = "2017-05-14 06:00")
plot(vp_6UTC, quantity = "ff")
```

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# Plot the ground speed (ff):
plot(vp_6UTC, quantity = "ff")
# Speeds at 1500 metre is approximately 12 m/s, confirming our earlier reading above of 12.5 m/s.
```

## Vertical and time integration of profiles

Often you will want to sum all the migrants across altitude, for example if you are interested in a single index of how many birds are migrating at a certain instant. There are at least two ways in which you can do that:

* by calculating the vertically integrated bird density (VID), which is _surface_ density as opposed to a _volume_ densities you have been plotting in the previous exercises: this number gives you how many migrants are aloft per square kilometer earth's surface (unit individuals/km$^{2}$), obtained by a vertical integration of the volume densities (unit individuals/km$^{3}$).

$$\textrm{VID}=\sum_{\textrm{height layers } i}{(\textrm{bird volume density})_i \times (\textrm{height layer thickness})_i}$$

* Note that the VID quantity doesn't depend on the speed of the migrants. A common measure that reflects both the density and speed of the migration is migration traffic rate (MTR). This is flux measure that gives you how many migrants are passing the radar station per unit of time and per unit of distance perpendicular to the migratory direction (unit individuals/km/hour).

$$\textrm{MTR}=\sum_{\textrm{height layers } i}{(\textrm{bird volume density})_i \times (\textrm{bird speed})_i \times (\textrm{height layer thickness})_i}$$

We will be using bioRad's `integrate_profile()` function to calculate these quantities:

```{r, eval=SHOW_ANSWERS}
# Let's continue with the vpts object created in the previous example.
# The vertically integrated quantities are calculated as follows:
my_vpi <- integrate_profile(my_vpts)
# The my_vpi object you created is a vpi class object, which is an acronym for "vertical profile integrated". It has its own plot method, which by default plots migration traffic rate (MTR):
plot(my_vpi)
# you can also plot vertically integrated densities (VID):
plot(my_vpi, quantity = "vid")
# the gray and white shading indicates day and night, which is calculated
# from the date and the radar position. You can also turn this off:
plot(my_vpi, night_shade = FALSE)
# plot the cumulative number of birds passing the radar, i.e. migration traffic (mt):
plot(my_vpi, quantity = "mt")
# execute `?plot.vpi` to open the help page listing all the options.
?plot.vpi
```
Both MTR, VID and MT depend on the assumed radar cross section (RCS) per bird. If you are unwilling/unable to specify RCS, alternatively you can use of closely related quantities that make no assumptions RCS:

```{r, eval=SHOW_ANSWERS}
# instead of vertically integrated density (VID), you can use vertically integrated reflectivity (VIR):
plot(my_vpi, quantity = "vir")
# instead of migration traffic rate (MTR), you can use the reflectivity traffic rate (RTR):
plot(my_vpi, quantity = "rtr")
# instead of migration traffic (MT), you can use the reflectivity traffic (RT):
plot(my_vpi, quantity = "rt")
```

VIR gives you the total cross-sectional area of air-borne targets per square kilometer of ground surface, whereas RTR gives you the total cross-sectional area of targets flying across a one kilometer line perpendicular to the migratory flow per hour.

## Manipulating vertical profile time series data
You may sometimes want to manipulate `vpts` data using other well-known data manipulations tools, like the R package dplyr.

```{r, eval=SHOW_ANSWERS}
# convert vpts object to a data.frame:
example_vpts |>
  as_tibble() -> my_df
my_df

# convert data.frame back to vpts object:
my_df |> as.vpts()

# as.vpts and as.data.frame (or as_tibble) work as each others inverse:
as.vpts(as.data.frame(example_vpts))
```

This allows for tidyverse style data manipulations. For example, let's filter out data by height and date using package dplyr and plot it, using a single command:
```{r, eval=SHOW_ANSWERS}
example_vpts |>
  # convert to data.frame, without adding sunrise/sunset columns
  as.data.frame(suntime=FALSE) |>
  # filter out heights layers above 1500 meter
  filter(height <= 1500) |>
  # filter out data from Sep 4 - Sep 9
  filter(datetime < as.POSIXct("2016-09-04",tz="UTC") |
         datetime > as.POSIXct("2016-09-07",tz="UTC")) |>
  # convert back to vpts object
  as.vpts() |>
  # regularize data on a regular time grid
  # (so we can visualize the data gap we created):
  regularize_vpts() |>
  # plot the vertical profile time series
  plot()
```

Or let us add a data column `high_speed` that indicates scatterers moving faster than 15 m/s, and plot that new data column in a profile plot:
```{r, eval=SHOW_ANSWERS}
example_vpts |> 
  # convert vpts object to a data.frame:
  as.data.frame() |>
  # add a data column that is TRUE (1) when the ground speed ff > 15 m/s,
  # and FALSE (0) if it is slower or equal to 15 m/s:
  mutate(high_speed = ff > 15) |>
  # convert back to vpts object:
  as.vpts() |>
  # plot the new quantity:
  plot(quantity="high_speed")
```

Note that compatibility with the tidyverse remains under development, and critical data columns of vpts objects (like height) should not be removed. 

## Inspecting precipitation signals in profiles

Precipitation is known to have a major influence on the timing and intensity of migration, therefore it is a useful skill to be able to inspect profiles for presence of precipitation.

Also, although automated bird quantification algorithms become more and more reliable, distinguishing precipitation from birds remains challenging for algorithms in specific cases. It is therefore important to have the skills to inspect suspicious profiles. That may help you to identify potential errors of the automated methods, and prevent your from overinterpreting the data.

An easy way of doing that is plotting the vertical profile of total reflectivity (quantity DBZH), which includes everything: birds, insects and precipitation. Precipitation often has higher reflectivities than birds, and also extends to much higher altitudes.

```{r, eval=SHOW_ANSWERS}
# load a time series for the KBGM radar in Binghamton, NY
my_vpts <- get_vpts("KBGM",as.Date(c("2026-04-16","2026-04-20")), source="birdcast")
# print the loaded vpts time series for this radar:
my_vpts
# plot the bird density over time:
plot(my_vpts, quantity = "dens", ylim=c(0,3000))
```

**Exercise 9:** Compare the above plot for bird density (quantity `dens`) with a profile plot for total reflectivity (quantity `DBZH`, showing birds and precipitation combined). Compare the two plots to visually identify periods and altitude layers with precipitation.

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# also show meteorological signals:
plot(my_vpts, quantity = "DBZH", ylim=c(0,3000))

# Periods with high reflectivities extending to high altitudes indicate precipitation,
# i.e. second half of the second night, and on and off during the fourth night.
```

We can automatically identify precipitation posthoc by identifying extensive vertical bands of high DBZH. This is where function `filter_precip()` comes in handy. It can be used as a postdoc filter for rain contaminations, however it will discard cases where birds and rain co-occur:

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# - `alt_max` specifies the maximum altitude to consider
# - `dbz` is the minimum reflectivity factor to consider for precipitation filtering
# - `range` specifies how much altitude range in meters should be filled with precip (reflectivity factor>dbz) in order for the profile to be filtered out.
my_vpts |>
  filter_precip(range=2000, dbz=15, alt_max=3000) |>
  plot(ylim=c(0,3000))
# Use with care, optimal parameters will depend on your use case!
```


# Range bias correction
In a previous section we have explored vertical integration of vertical profiles (`vp`'s). In this paragraph we will generalize vertical integration to entire radar images (`ppi`'s). This requires us to account for the changing beam shape of the radar with range.

First, let's examine the beam shape of the lowest elevation scan of the radar, which is typically around 0.5 degrees.

```{r, eval=SHOW_ANSWERS}
# define ranges from 0 to 2500000 meter (250 km), in steps of 100 m:
range <- seq(0, 250000, 100)

# plot the beam height of the 0.5 degree elevation beam:
plot(range, beam_height(range, 0.5), ylab = "beam height [m]", xlab = "range [m]", type='l')

# let's add the lower and upper altitude of the beam, as determined by the beam width:
points(range, beam_height(range, 0.5)-beam_width(range)/2, type='l',lty=3)
points(range, beam_height(range, 0.5)+beam_width(range)/2, type='l',lty=3)
```

We will start with downloading a polar volume and processing it into a profile:

## Processing a polar volume into a profile
```{r, eval=SHOW_ANSWERS, warning=FALSE, results='hide'}
# download a polar volume for the KBRO radar in Brownsville, TX
my_pvol <- get_pvol("KBRO",as.POSIXct("2017-05-14 05:50:00"))
# calculate the profile, correlation-coefficient is used automatically to remove precipitation:
# we calculate 60 layers of 50 meter width, so up to 60*50=3000 m.
my_vp <- calculate_vp(my_pvol, n_layer=60, h_layer=50)
# also clean the pvol object for precipitation using mistnet:
my_pvol <- apply_mistnet(my_pvol)
# set all the DBZH pixels identified by mistnet as precipitation to NA:
my_pvol_clean <- calculate_param(my_pvol, DBZH = ifelse(CELL == 2, NA, DBZH))
```

**Exercise 10:** Plot the bird density in the vertical profile you just estimated and compare it with the plots above of the beam height and width of the lowest radar beam. At which approximate range do you expect the radar will no longer be able to resolve the altitude distribution of the migratory birds. And at which range will the radar start to overshoot the migration layer entirely?

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# plot the profile
plot(my_vp)

# Answer:
# * The vertical profile shows altitude bands of about 300 meter width
#   We therefore expect the radar to start having difficulty to resolve this profile
#   when the beam shape becomes broader, so less than 50 km
#   (this is why we typically use 35 km as the maximum range to use in vertical profile estimation.)
# * the birds fly up to about 2km in this profile.
#   At about 200000 m (200 km) the lower end of the beam no longer overlaps with the migration layer
#   Therefore, if we birds are flying according to the same altitude distribution everywhere,
#   we expect the radar to become fully blind for birds beyond 200 km
```

To correct for the decreasing ability of the radar to resolve the altitude distributions of birds with range, we will make one important (and likely imperfect!) assumption:

**We assume that all birds within the image are distributed vertically according to the same relative vertical profile.**

This assumptions simplifies the problem, and allows us to estimate the spatial distribution of the birds, as we will explore in the next paragraph:

## Range bias correction and vertical integration on a map
```{r, eval=SHOW_ANSWERS, warning=FALSE}
# We will use this helper function to create a base grid centered around the radar
# distance is the maximum distance from the radar for the grid, res is
# the pixels size, both in meters
define_base_grid <- function(pvol, distance=100000, res=1000, shape_only=FALSE){
  # define the base grid area
  base_grid_area <-
    # define a point at the location of the radar in WGS84 coordinates
    sf::st_sfc(sf::st_point(c(pvol$attributes$where$lon, pvol$attributes$where$lat)), crs = 4326) |>
    # convert to Albers Equal-Area Conic projection, which is good for preserving the relative sizes
    # of areas in the contiguous USA.
    sf::st_transform(crs = 5070) |>
    # create a 100 km buffer radius around the radar location
    sf::st_buffer(dist = distance) |>
    # convert to sf dataframe
    sf::st_as_sf()
  # if shape_only is TRUE, only return the circular shape:
  if(shape_only) return(base_grid_area)
  # otherwise, return a full terra raster:
  # create a raster for the base grid area
  terra::rast(
    ext = terra::ext(base_grid_area),  # extent from the circular area
    resolution = res,          # 1000m resolution
    crs = terra::crs(base_grid_area)   # same CRS as the circular area
  )
}

# define the base grid
base_grid <- define_base_grid(my_pvol_clean)
# Calculated the vertically integrated PPI
my_ppi_integrated <- integrate_to_ppi(pvol=my_pvol_clean,vp=my_vp, raster=base_grid)
```

Next, we will compare the PPI for the 1.0 degree sweep and the vertically integrated PPI, to highlight the effect of the range correction and vertical integration. We will plot the log-transform of VID for a more direct comparison with the reflectivity factor DBZH, which is also logarithmic in bird density.
```{r, eval=SHOW_ANSWERS, warning=FALSE}
# Project the 1 degree sweep on the same basegrid:
my_pvol_clean |>
  get_scan(elev = 1.0) |>
  project_as_ppi(raster=base_grid) ->
  my_ppi
# let's take the logarithm of the vertically integrated density, so
# we can compare it more directly to DBZH, which is also a logarithmic quantity:
my_ppi_integrated <- calculate_param(my_ppi_integrated, logVID=log(VID))
# plot both ppi's:
plot(my_ppi)
plot(my_ppi_integrated, param="logVID", zlim=c(-10,10))
```


**Exercise 11:** Explain why we see concentric rings in the 1.0 degree sweep, and not in the vertically integrated PPI.

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS}
# The altitude distribution in this case shows two migration layers.
# In the 1.5 PPI scan these show up as two concentric rings of density.
# Since in a conventional PPI a larger distance from the radar also
# implies a higher altitude, we see the rings showing up at the
# ranges where the beam intersects each of the migration layers.
#
# The vertically integrated PPI is more straightforward to interpret
# Here we correct for aforementioned beam effects and integrate the
# density over altitude, giving a more realistic reconstruction of the
# spatial distribution of migrants.
```


## Stopover mapping
Birds taking off tend to still be closely associated with ground level. In areas with topography, the common assumption is that birds will still take off simultaneously, resulting in an altitude distribution
during the take-off phase that is best estimated relative to ground level. 

To calculate a profile referenced to ground level, we first define a base grid and download digital elevation information:
```{r, eval=SHOW_ANSWERS}
# Download a polar volume for KCCX radar
# This in the Appalachian mountains of Central Pennsylvania
pvol <- getRad::get_pvol("KCCX", as.POSIXct("2013-10-08 23:24:30", tz="UTC"))
# This polar volume is collected 45 mins after local sunset.
# Birds are still in the process of take-off during this time.
pvol$datetime - sunset(pvol$datetime, pvol$geo$lon, pvol$geo$lat)
# define the base grid area:
base_grid_area <- define_base_grid(pvol, shape_only=TRUE)
# download digital elevation grid (z defines resolution), and name the data "HGHT"
base_grid_ground <- elevatr::get_elev_raster(base_grid_area, z=7, clip = "locations")
names(base_grid_ground) = "HGHT"
# Plot the topography as a reference:
plot(rast(base_grid_ground))
# add ground height data as a scan parameter to the polar volume:
pvol_ground <- add_param(pvol, raster=base_grid_ground, param="HGHT")
# verify that ground height information is now added to each elevation scan:
plot(get_scan(pvol_ground,.5), param="HGHT")
```

Next, we can calculate a vertical profile relative to ground level:
```{r, eval=SHOW_ANSWERS}
# calculate a profile relative to ground level
vp_ground <- calculate_vp(pvol_ground, height_reference = "ground")
# calculate a range-corrected image, assuming birds maintain a constant
# altitude profile relative to ground level:
my_ppi_ground <- integrate_to_ppi(pvol=pvol_ground,vp=vp_ground, raster=base_grid_ground, height_reference="ground")
# Mask out areas with very strong range correction that are unreliable.
# In this case we require eta_sum_to_VIR<=8:
my_ppi_ground <- calculate_param(my_ppi_ground, VIR=ifelse(eta_sum_to_VIR>8,NA,VIR))
# Plot the result, using viridis color scale:
plot(my_ppi_ground, zlim=c(0,2000)) + viridis::scale_fill_viridis(name = "VIR")
```

**Exercise 12:** Create the same stopover map, but instead of assuming birds maintain
a constant altitude distribution relative to ground level, assume they maintain a
constant altitude distribution relative to sea level. Explain observed differences
in the two maps.

```{r, echo=SHOW_ANSWERS, eval=SHOW_ANSWERS, warning=FALSE}
# calculate the profile relative to sea level
vp_sea <- calculate_vp(pvol_ground, height_reference = "sea")
# inspect the profile differences
plot(vp_sea)
plot(vp_ground)
# range correction relative to sea level:
my_ppi_sea <- integrate_to_ppi(pvol=pvol_ground,vp=vp_sea, height_reference="sea", raster=base_grid_ground)
# mask out areas with very strong range correction
my_ppi_sea <- calculate_param(my_ppi_sea, VIR=ifelse(eta_sum_to_VIR>8,NA,VIR))
# plot the result, using viridis color scale:
# More departure is inferred in low elevation valleys below the antenna height
# for the ground referenced range-correction:
plot(my_ppi_ground, zlim=c(0,2000)) + viridis::scale_fill_viridis(name = "VIR")
plot(my_ppi_sea, zlim=c(0,2000)) + viridis::scale_fill_viridis(name = "VIR")
```

# Processing many polar volumes

## Accessing radar data

* US NEXRAD polar volume data can be accessed in the [Amazon cloud](https://s3.amazonaws.com/noaa-nexrad-level2/index.html). Use function `download_pvolfiles()` for local downloads
* European radar data can be accessed at https://aloftdata.eu. These are processed vertical profiles, the full polar volume data are not openly available for most countries. Data are available for single profiles (vp) in ODIM h5 format, or as monthly vertical profile time series (vpts) in [VPTS CSV](https://aloftdata.eu/vpts-csv/) format.
* Use function `get_vpts()` to download VPTS data from public repositories (aloft, birdcast, dark ecology).
* Use function `get_pvol()` to download polar volume data from public repositories (aloft, NEXRAD).

The names of the radars in the networks can be found here:

* US: [NEXRAD network](http://weather.gov/nl2/NEXRADView)
* Europe: [OPERA database](http://eumetnet.eu/wp-content/themes/aeron-child/observations-programme/current-activities/opera/database/OPERA_Database/index.html)

Useful sites for inspecting pre-made movies of the US composite are <https://birdcast.info/migration-tools/live-migration-maps/> and <https://www.pauljhurtado.com/US_Composite_Radar/>.


## Processing multiple polar volumes

This section contains an example for processing a directory of polar volumes into profiles:

It tends to be faster to first download a batch of files before processing them. Therefore we first we download files locally, and prepare an output directory for storing the processed profiles. 
```{r, eval=SHOW_ANSWERS,results='hide',warning=FALSE}
# First we download more data, for a total of one additional hour for the same radar:
# create a directory to store polar volumes (pvol data) and processed profiles (vp data)
dir.create("./data_pvol")
download_pvolfiles(date_min=as.POSIXct("2026-05-18 00:45:00"), date_max=as.POSIXct("2026-05-18 02:30:00"), radar="KBGM", directory="./data_pvol")
# We will process all the polar volume files downloaded so far:
my_pvolfiles <- list.files("./data_pvol", recursive = TRUE, full.names = TRUE, pattern="KBGM")
my_pvolfiles
# create output directory for processed profiles
outputdir <- "./data_vp"
dir.create(outputdir)
```

We will use the following custom settings for processing:
* We will use MistNet to screen out precipitation
* we will calculate profile layers of 50 meter width
* we will calculate 60 profile layers, so up to 50*60=3000 meter altitude
* we will store the profiles in [VPTS CSV](https://aloftdata.eu/vpts-csv/) format, by specifying an output file with extension .csv (this file format is fast for reading data back in later)

Note that we enclose the `calculate_vp()` function in `tryCatch()` to keep going after errors with specific files
```{r, eval=FALSE, echo=FALSE}
# let's loop over the files and generate profiles
start=Sys.time()
for (file_in in my_pvolfiles) {
  # generate the output filename for the input file
  # .csv extension produces file in VPTS CSV format (human readable text)
  # (Without .csv extension you get ODIM HDF5 format)
  file_out <- paste(outputdir, "/", basename(file_in), "_vp.csv", sep = "")
  # generate the profile, and write it as a VPTS CSV file (for later reference)
  # we enclose the calculate_vp statement in tryCatch to keep going after errors with specific files
  vp <- tryCatch(calculate_vp(file_in, file_out, mistnet=TRUE, h_layer=50, n_layer=60), error = function(e) NULL)
}
end=Sys.time()
# calculate the processing time that has passed:
end-start
```

Having generated the profiles, we can read them into R:

```{r, eval=SHOW_ANSWERS}
# we assume outputdir contains the path to the directory with processed profiles
my_vpfiles <- list.files(outputdir, full.names = TRUE, pattern="KBGM")
# print them
my_vpfiles
# read them into a vpts object:
my_vpts <- read_vpts(my_vpfiles)
```

You can now continue with visualizing and post-processing as we did earlier:

```{r, eval=SHOW_ANSWERS}
# plot them between 0 - 3 km altitude:
plot(my_vpts, ylim = c(0, 3000))
# let's visualize rain by plotting all aerial reflectivity:
plot(my_vpts, ylim = c(0, 3000), quantity="DBZH")
```
Note that we enclose the `calculate_vp()` function in `tryCatch()` to keep going after errors with specific files

## Parallel processing
We may use one of the parallelization packages in R to further speed up our processing. We will use `mclapply()` from package `parallel`. First we wrap up the processing statements in a function, so we can make a single call to a single file. We will disable MistNet, as this deep-learning model does not parallelize well on a cpu machine.
```{r, eval=SHOW_ANSWERS}
process_file <- function(file_in){
  # construct output filename from input filename
  file_out <- paste(outputdir, "/", basename(file_in), "_vp.csv", sep = "")
  # run calculate_vp()
  vp <- tryCatch(calculate_vp(file_in, file_out, mistnet = FALSE, h_layer=50, n_layer=60), error = function(e) NULL)
  if(is.vp(vp)){
    # return TRUE if we calculated a profile
    return(TRUE)
  } else{
    # return FALSE if we did not
    return(FALSE)
  }
}

# To process a file, we can now run
process_file(my_pvolfiles[1])
```
Next, we use `mclapply()` to do the parallel processing for all files:
```{r, eval=FALSE}
# load the parallel library
library(parallel)
# detect how many cores we can use. We will keep 2 cores for other tasks and use the rest for processing.
number_of_cores = detectCores() - 2
# Available cores:
number_of_cores
# let's loop over the files and generate profiles
start=Sys.time()
mclapply(my_pvolfiles, process_file, mc.cores = number_of_cores)
end=Sys.time()
# calculate the processing time that has passed:
end-start
```

# Further analysis outside bioRad
In many cases you will want to convert bioRad's objects into a convenient form for your own further analyses. To convert bioRad objects to a simple `data.frame`:

* vertical profile (`vp`): as.data.frame(`your_vp_object`)
* vertical profile time series (`vpts`): as.data.frame(`your_vpts_object`)
* use function `as.vpts()` to convert a vpts data.frame back to a vpts object.
* integrated vertical profile time series (`vpi`): is already a `data.frame`

Converting polar scans (`scan` objects) to common spatial formats:

* use `scan_to_raster()` to convert to a `RasterBrick` object (package raster)
* use `scan_to_spatial()` to convert to a `SpatialPointsDataFrame` object (package sp)

Converting PPI (`ppi` objects) to common spatial formats:
* access the data slot (`my_ppi$data`) to extract a `SpatialGridDataFrame` object (package sp)

# Troubleshooting

## Cannot find proj.db
Message `pj_obj_create: Cannot find proj.db` shows up when using pre-compiled binaries generated by CRAN. The message is harmless but annoying. You can silence it by pointing environment variable `PROJ_LIB` to the `share/proj` folder of the proj system library.

On Mac OSX you can install the proj library from a terminal using [Homebrew](https://brew.sh/):

```
brew install proj
```

Next, point the `PROJ_LIB` environment variable to the location where your library is installed (may vary by system), for example Homebrew tends to install libraries in `/opt/homebrew/Cellar/`:

```{r, eval=SHOW_ANSWERS}
Sys.setenv(PROJ_LIB = "/opt/homebrew/Cellar/proj/9.7.1/share/proj/")
```
